---
title: "EDA"
format: html
toc: TRUE
---

```{r}
#| include: FALSE

library(tidyverse)
```

## Introduction

Hmmmm.

Read in file

```{r}
diabetes <- read_csv(file = "diabetes_binary_health_indicators_BRFSS2015.csv")
```

```{r}
diabetes
```

According to the [documentation](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/) of the dataset in kaggle:

- Diabetes_binary: 0 = no diabetes, 1 = prediabetes or diabetes
- Binary fields indicating 1 for presence or 0 for absence of the specified condition 
    + HighBP, HighChol, Stroke, HeartDiseaseorAttack
    + DiffWalk: difficulty walking or climbing stairs
- Binary fields that need additional description to understand what it is representing
    + CholCheck: Cholesterol check in 5 years?
    + Smoker: Have you smoked at least 100 cigarettes in your entire lifetime?
    + PhysActivity: physical activity in the past 30 days - not including job
    + Fruits: Consume fruit 1 or more times per day
    + Veggies: Consume veggies 1 or more times per day
    + HeavyAlcoholConsump: Adult men >= 14 drinks per week, Adult women >= 7 drinks per week
    + AnyHealthcare: Has any kind of healthcare/insurance coverage
    + NoDocbcCost: Any time in the last 12 months respondent needed to see a doctor, but did not due to cost?
- Numeric fields
    + BMI
    + MentHlth: days of poor mental health in the last 30 days
    + PhysHlth: days of physical illness or injury in the last 30 days
- Numeric fields, but represent levels. Documentation provided a link to the CDC codebook [here](https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf).
    + GenHlth: self-rating of personal health from 1 = excellent to 5 = poor
    + Age: 1 = 18-24, 2 = 25-29, 3 = 30-34, 4 = 35-39, 5 = 40-44, 6 = 45-49, 7 = 50-54, 8 = 55-59, 9 = 60-64, 10 = 65-69, 11 = 70-74, 12 = 75-79, 13 = 80+
    + Education: 1 = Never attended, or only kindergarten, 2 = Grades 1-8 (Elementary), 3 = Grades 9-11 (some high school), 4 = Grades 12 or GED (High school graduate), 5 = College 1-3 years (Some college or technical school), 6 = College 4 years or more (College graduate)
    + Income: 1 = Less than $10,000, 2 = $10,000-14,999, 3 = $15,000-19,999, 4 = $20,000-24,999, 5 = $25,000-34,999, 6 = $35,000-$49,999, 7 = $50,000-74,999, 8 = $75,000+
    + Sex: 0 = Female, 1 = Male


```{r}
psych::describe(diabetes)
```

- The max number for BMI looks kind of surprising, at 98. Since BMI > 30 is obese, I think it makes sense to look at the distribution of the values for this variable to see if it looks reasonable for a sample that should somewhat represent the general population. The median of 27 and relatively low standard deviation of 6.61 show it's most likely the really high values are outliers, but it would be good to get a more thorough look at summary statistics.
- All other values appear consistent with what the documentation describes. There are a lot of binary variables, and then there are a collection of variables with coded numeric values that need to be translated to a descriptive label.

### Data Cleaning

Recast data columns

```{r}
diabetes <- diabetes |>
  mutate(Diabetes_binary = as.factor(Diabetes_binary),
         HighBP = as.factor(HighBP),
         HighChol = as.factor(HighChol),
         CholCheck = as.factor(CholCheck),
         Smoker = as.factor(Smoker),
         Stroke = as.factor(Stroke),
         HeartDiseaseorAttack = as.factor(HeartDiseaseorAttack),
         PhysActivity = as.factor(PhysActivity),
         Fruits = as.factor(Fruits),
         Veggies = as.factor(Veggies),
         HvyAlcoholConsump = as.factor(HvyAlcoholConsump),
         AnyHealthcare = as.factor(AnyHealthcare),
         NoDocbcCost = as.factor(NoDocbcCost),
         DiffWalk = as.factor(DiffWalk),
         GenHlth = factor(GenHlth,
                          levels = 1:5,
                          labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),
         Sex = factor(Sex, levels = 0:1, labels = c("Female", "Male")),
         Age = factor(Age,
                      levels = 1:13,
                      labels = c("18 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", 
                                 "45 to 49", "50 to 54", "55 to 59", "60 to 64", "65 to 69", 
                                 "70 to 74", "75 to 79", "80 and over")),
         Education = factor(Education,
                            levels = 1:6,
                            labels = c("Never attended/Only Kindergarten", "Elementary", 
                                       "Some high school", "High school graduate", 
                                       "Some college/tech school", "College graduate")),
         Income = factor(Income,
                         levels = 1:8,
                         labels = c("Less than $10,000", "$10,000 to $14,999", "$15,000 to $19,999", 
                                    "$20,000 to $24,999", "$25,000 to $34,999", "$35,000 to $49,999",
                                    "$50,000 to $74,999", "$75,000 or greater"))
         )
```

Make sure the factors that had coded values changed to descriptive ones look right

```{r}
diabetes |>
  group_by(Sex) |>
  summarize(count = n())
```

```{r}
diabetes |>
  group_by(Age) |>
  summarize(count = n())
```

```{r}
diabetes |>
  group_by(GenHlth) |>
  summarize(count = n())
```

```{r}
diabetes |>
  group_by(Education) |>
  summarize(count = n())
```

```{r}
diabetes |>
  group_by(Income) |>
  summarize(count = n())
```

Check for missing values

```{r}
colSums(is.na(diabetes))
```
### Summary Statistics

Going to look at the five number summary for BMI
```{r}
diabetes |>
  group_by(Diabetes_binary) |>
  summarize(min = min(BMI),
            q1 = quantile(BMI, probs = 0.25),
            median = median(BMI),
            q3 = quantile(BMI, probs = 0.75),
            max = max(BMI))
```

This looks reasonable, but I'll still do a density plot to make sure. Mainly to make sure that the values in the lower 25th and the upper 75th percentiles aren't all grouped at the extremes, but instead spread out a bit. If everything above 35 or so are all 98's for instance, I'd be concerned that 98 is actually a coded value for something rather than a true value (e.g. what if the true values are capped at 35, and anything exceeding it was coded as 98). If they're spread out, then I'd be more comfortable assuming they're valid outliers.
   
   
```{r}
diabetes |>
  group_by(Age, Diabetes_binary) |>
  summarize(across(where(is.numeric),
            mean,
            .names = "mean_{col}"))

```


#### Correlation to Diabetes status

```{r}

# see if there are correlations between Diabetes status and all other columns
correlation <- diabetes |>
  lapply(as.numeric) |>
  as.data.frame() |>
  summarise(across(everything(), 
                   \(x) list(cor.test(x, as.numeric(diabetes$Diabetes_binary))))) 

# transform the correlation statistics from list-column to a flat numeric column
corr_col <- correlation |>
  map(1) |>                 # access first element (a list) of each list-column
  map("estimate") |>        # pull just the estimate element for each inner list
  bind_rows()               # compress into a simple column

tibble(Variable = names(correlation)) |>
  bind_cols(corr_col) |>
  filter(abs(cor) > 0.02, cor < 1.0) |>
  arrange(desc(abs(cor)))
  
```

Might be good to focus on the top 5 variables in the above list as predictors. These variables are consistent with the fact that diabetes is associated with heart disease, as discussed in the [documentation](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data) for the dataset. Difficulty walking may be related to neuropathy, which is also a complication of diabetes, I think related to blood flow, which is compromised by diabetes. Problems with circulation can also impact general health by reducing the effectiveness of the immune response (immune cells will have a harder time reaching sites of infection and damage if there are issues with blood flow). Therefore having some relationship between GenHlth and diabetes status is reasonable.

Look at the correlations between those 5 predictors and all other variables to see if they have associations with other variables that might also be interesting to look at.

```{r}
correlation_2 <- diabetes |>
  map(as.numeric) |>
  as_tibble() |>
  cor()

# just look at top 5
correlation_2[,c("GenHlth", "HighBP", "DiffWalk", "BMI", "HighChol")]
```

Nothing particularly highly correlated, just some weak associations between those "top 5" variables already identified, plus some association with GenHlth and DiffWalk/PhysHlth/Ment, and also some very weak correlation with heart attack history and the "top 5".

#### Contingency Tables

Using the above correlations as a guide, look at contingency tables of the correlation values that are the farthest from 0 (although none are particularly high).

```{r}
cont_GenHlth <- diabetes |>
  group_by(GenHlth, Diabetes_binary) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Diabetes_binary, 
              values_from = count, 
              names_prefix = "Diabetes_") |>
  mutate(total = Diabetes_0 + Diabetes_1,
         Diabetes_0_pct = Diabetes_0 / total,
         Diabetes_1_pct = Diabetes_1 / total)

cont_GenHlth
```

```{r}
cont_HighBP <- diabetes |>
  group_by(HighBP, Diabetes_binary) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Diabetes_binary, 
              values_from = count, 
              names_prefix = "Diabetes_") |>
  mutate(total = Diabetes_0 + Diabetes_1,
         Diabetes_0_pct = Diabetes_0 / total,
         Diabetes_1_pct = Diabetes_1 / total)

cont_HighBP
```

```{r}
cont_DiffWalk <- diabetes |>
  group_by(DiffWalk, Diabetes_binary) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Diabetes_binary, 
              values_from = count, 
              names_prefix = "Diabetes_") |>
  mutate(total = Diabetes_0 + Diabetes_1,
         Diabetes_0_pct = Diabetes_0 / total,
         Diabetes_1_pct = Diabetes_1 / total)

cont_DiffWalk
```

```{r}
cont_BMI <- diabetes |>
  group_by(Diabetes_binary) |>
  summarize(count = n(),
            mean_BMI = mean(BMI))

cont_BMI
```

```{r}
cont_HighChol <- diabetes |>
  group_by(HighChol, Diabetes_binary) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Diabetes_binary, 
              values_from = count, 
              names_prefix = "Diabetes_") |>
  mutate(total = Diabetes_0 + Diabetes_1,
         Diabetes_0_pct = Diabetes_0 / total,
         Diabetes_1_pct = Diabetes_1 / total)

cont_HighChol
```

### Plots

Bar chart showing how a higher proportion of respondents reporting Good, Fair, or Poor health have a higher rate of diabetes/prediabetes than respondents reporting Very Good or Excellent health.

```{r}
g <- ggplot(data = diabetes, aes(x = GenHlth))
g + geom_bar(aes(fill = Diabetes_binary)) 

```

Because there was some correlation between the diabetes status and GenHlth, and there was also correlation between GenHelth and the days of physical and/or mental health difficulties, I thought it would be cool to put them on a contour map to see if there were patterns for diabetes status vs Physical and Mental health days.

When I initially plotted the original data points for physical and mental health days, which had a range of 0-30, there were many spots on the plot that were blank because there were no respondents that gave that particular combination of days. I set up a rounded column for each of these variables which rounded the number down to the nearest 5 (so 0-4 = 0, 5-9 = 5, 10-14 = 10, etc.). 

I also added in new column just for the plot which gives the proportion in each of those combinations had diabetes or prediabetes. This was used as the z-value for the plot. 

```{r}
# note: Diabetes_binary is converted to character first since it's a factor, and converting directly to numeric will cause the number to take on its factor level value (i.e. it would change 0 and 1 to 1 and 2)
diabetes_rollup <- diabetes |>
  mutate(PhysHlth_rnd = floor(PhysHlth/5)*5,
         MentHlth_rnd = floor(MentHlth/5)*5) |>
  group_by(PhysHlth_rnd, MentHlth_rnd) |>
  summarise(Diabetes_prop = sum(as.numeric(as.character(Diabetes_binary)))/n())

h <- ggplot(data = diabetes_rollup, aes(x = PhysHlth_rnd, y = MentHlth_rnd)) 
h + geom_contour_filled(aes(z = Diabetes_prop)) +
  labs(x = "Days with Physical Health Difficulties",
       y = "Days with Mental Health Difficulties",
       fill = "Proportion \nwith diabetes/ \nprediabetes") 

```

This shows a cool pattern where the more days that a person reported poor health days, the higher the proportion of diabetes/prediabetes. This effect looks more related to physical health issues when there are fewer than 10-15 days with those types of issues, but as the number of days gets higher, the people with more days of mental health struggles start to have higher rates of diabetes/prediabetes. Anecdotally, this seems consistent with the assumption that a person with these conditions who reports many days of the month with physical health burdens may start to struggle with mental health as well.


```{r}
diabetes |>
  distinct(PhysHlth, MentHlth) |>
  arrange(PhysHlth, MentHlth) |>
  mutate(PhysHlth_rnd = floor(PhysHlth/5)*5,
         Mentlth_rnd = floor(MentHlth/5)*5)
```


Boxplot for BMI to see the distribution of values

```{r}
g <- ggplot(data = diabetes, aes(x = Diabetes_binary, y = BMI, fill = Diabetes_binary))

g + geom_boxplot()
```
Interesting to note that the median BMI is higher in the Has Prediabetes/Diabetes group than the other. But both groups have a large number of BMI's outside the middle 50%, which is interesting.

```{r}
g <- ggplot(data = diabetes, aes(x = BMI, fill = Diabetes_binary))

g + geom_density(alpha = 0.5) +
  geom_vline(aes(xintercept = median(BMI[Diabetes_binary == 0])), 
             color = "darkgreen", 
             linetype = "dashed") +
  geom_vline(aes(xintercept = median(BMI[Diabetes_binary == 1])), 
             color = "navy", 
             linetype = "dashed") +
  scale_fill_manual(values = c("paleturquoise", "peachpuff")) +
  labs(x = "BMI", y = "Proportion")
```

Density plot provides another view showing that the prediabetes/diabetes group has higher BMI on average, and we can also see that although there's a long tail on the right side, there aren't a huge number of points in the extremes relative to the size of the data set.
 



[Go to the Modeling Page](Modeling.html)